Free surface computations in mantle convection models

### 2015-07-30 free-surface-paper.tex
Adjust figures
Reorganize things, make reference state clear

Here we analyze the problem of the sloshing instability.  We demonstrate the nature of the instability, 
as well as clarifying the effect of quasi-implicit stabilization.

The analysis allows us to propose a new time marching scheme for free surface computations with 
good stability properties. It is an explicit method which requires no modification to the system matrix,
which makes it extremely simple to implement.

@@

@@

We begin with the homogeneous Stokes equations for incompressible creeping flow:
\begin{equation}
\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p = 0
\label{eq:stokes}
\end{equation}

\begin{equation}
\nabla \cdot \mathbf{u} = 0
\label{eq:incompressible}
\end{equation}
where $\mathbf{u}$ is the velocity, $p$ is the pressure, $\eta$ is the viscosity, and 
$\varepsilon(\mathbf{u}) = \frac{1}{2}(\nabla \mathbf{u} + (\nabla \mathbf{u} )^T )$ is the strain-rate tensor.

@@

\begin{equation}
-\int_\Omega 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u} ) + \int_\Omega p \nabla \cdot \mathbf{w} 
+ \int_{\partial \Omega_N} \mathbf{w} \cdot \mathbf{T} \cdot \mathbf{n} = 0 
\label{eq:weak_stokes}
\end{equation}

\begin{equation}
\int_\Omega q \nabla \cdot \mathbf{u} = 0
\label{eq:weak_incompressible}
\end{equation}
where $\mathbf{w}$ and $q$ are suitably chosen test functions and the integrals over 
$\Omega$ and $\partial \Omega_N$ are over the volume of the domain and the part of the boundary with stress boundary conditions, respectively.
$\mathbf{T}$ is the stress tensor, defined by
\begin{equation}
\mathbf{T} = 2 \eta \varepsilon(\mathbf{u}) - p \mathbf{I}
\end{equation}
The integral over the surface in Equation~\eqref{eq:weak_stokes} accounts for boundary stresses, 
which should be zero when evaluated on a true free surface.

Rather than analyzing finite deformation to the free surface (a nonlinear problem),
we will make the analytically useful approximation of small deformations about the hydrostatic 
reference surface and analyze their stability.
We will therefore  evaluate the integrals in Equation~\eqref{eq:weak_stokes} 
over the hydrostatic reference surface and introduce a temporary auxilliary variable $\zeta$ which 
represents the (small) topography of the free surface relative to that reference surface.
The Lagrangian equation for time evolution of the free surface is then
\begin{equation}
\frac{d \zeta}{dt} = \mathbf{u \cdot n}
\label{eq:surface_evolution}
\end{equation}
where $\mathbf{u}$ is evaluated at the free surface.
The stress on this reference surface can be approximated by the hydrostatic pressure
\begin{equation}
\mathbf{T} \approx \rho g \zeta \mathbf{I}
\label{eq:hydrostatic}
\end{equation}
where $\rho$ is the density and $g$ is gravity. Equation~\eqref{eq:weak_stokes} then becomes

\begin{equation}
-\int_\Omega 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u} ) + \int_\Omega p \nabla \cdot \mathbf{w} 
+ \int_{\partial \Omega_N} \rho g \zeta  \mathbf{w} \cdot \mathbf{n} = 0 
\end{equation}

We would like to analyze the time evolution of the normal modes of this system: each mode 
is the relaxation of topography with a characteristic relaxation time.  
We denote the normal modes by $\left[ \mathbf{u}_i, p_i, \zeta_i \right]^T$, with
relaxation times $\tau_i$, where the subscript corresponds to the $i$th normal mode.

@@

\begin{equation}
\frac {d}{d t} \zeta_i(\mathbf{x},t) = \frac{d}{dt} \zeta_i(\mathbf{x})e^{-t/\tau_i} = -\frac{\zeta_i(\mathbf{x},t)}{\tau_i} = \mathbf{u}_i \cdot \mathbf{n}
\end{equation}
This can then be used to eliminate $\zeta$ from the Stokes system:
\begin{equation}
-\int_\Omega 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u}_i ) + \int_\Omega p_i \nabla \cdot \mathbf{w} 
= \tau_i \int_{\partial \Omega_N} \rho g (\mathbf{u}_i \cdot \mathbf{n} ) (\mathbf{w} \cdot \mathbf{n})
\label{eq:weak_eigen}
\end{equation}
When these equations are discretized \citep[e.g.][]{kronbichler2012high} we get
\begin{equation}
\begin{bmatrix}
A & B^T \\
B & 0 \\
\end{bmatrix}
\begin{bmatrix}
\mathbf{u}^n_i \\
p^n_i
\end{bmatrix}
=
\tau_i
\begin{bmatrix}
M & 0 \\
0 & 0
\end{bmatrix}
\begin{bmatrix}
\mathbf{u}^n_i \\
p^n_i
\end{bmatrix}
\label{eq:generalized_eigenvalue}
\end{equation}
where $\mathbf{u}^n_i$ and $p^n_i$ are finite-dimensional representations of $\mathbf{u}_i$ and $p_i$,
and $M$ is the discretization of the bilinear form on the right-hand side of Equation~\eqref{eq:weak_eigen}.

@@

\begin{equation}
\int_{\Omega + \Delta \Omega} \rho  \mathbf{w} \cdot \mathbf{g} \approx
\int_{\Omega} \rho  \mathbf{w} \cdot \mathbf{g} + \theta \Delta t \int_{\partial \Omega} \rho ( \mathbf{w \cdot g}) (\mathbf{u \cdot n} )
\label{eq:kmm}
\end{equation}
The volume integral is the same as that of the unstabilized problem, but now there is a surface integral.
The surface integral has the form of a velocity-dependent surface stress pushing down on the 
domain, and can be thought of as an artificial viscous damping of the surface.
Since it depends on the velocity, it can no longer go on the right-hand side of the 
equation, and must be added as a stabilization term in the system matrix.  
Empirically it has been found to be successful at damping instabilities \citep{kaus2010stabilization, quinquis2011role, crameri2012comparison}.

Indeed, using the formalism in Section~\ref{sec:eigenvalue} we can see the effect of this 
stabilization term. In the reference state we note that $\mathbf{g} = g \mathbf{n}$, which
allows us to write the stabilization term as
\begin{equation}
\theta \Delta t \int_{\partial \Omega} \rho g ( \mathbf{w \cdot n}) (\mathbf{u \cdot n} )
\end{equation}

The integral here is precisely the same as that on the right hand side of Equation~\eqref{eq:weak_eigen}, which was discretized as the matrix $M$.

@@

\subsection{Stability of the scheme}
The parameter $\tau^*$ sets how quickly $\mathbf{u}$ decays in Equation~\eqref{eq:velocity_decay}, and a good 
choice is crucial for accuracy and stability. A shorter decay time corresponds to more stabilization,
but if it becomes too short, the surface velocity will become overdamped. In general, we will want 
to choose $\tau^*$ so that it is as close as possible to the least stable mode, or $\tau_{\mathrm{max}}$.

@@

The stability of this scheme is determined by the choice of $\tau^*$.  
It has a region of unconditional stability, where
\begin{equation}
\tau^* \le 2 \tau_i
\end{equation}
Otherwise, the scheme is conditionally stable, with the timestep limited by 
\begin{equation}
\Delta t \le -\tau^* \log \left(1 - 2 \frac{\tau_i}{\tau^*} \right)
\end{equation}
The stability region is plotted in Figure \ref{fig:stability}.

\begin{figure*}
\includegraphics[width=0.9\textwidth]{figures/stability.pdf}
\label{fig:stability}
\caption{Left: stability region for the nonstandard finite difference scheme (green). On the x-axis is the value of the stabilization time, in units of the minimum relaxation time $\tau_{\mathrm{min}}$.  On the y-axis is the value of the timestep, also in units of the minimum relaxation time. For $\tau^*\le 2 \tau_{\mathrm{min}}$ the nonstandard finite difference scheme is unconditionally stable.  Right: stability region for the quasi-implicit scheme (green).  Again, the y axis is in units of the minumum relaxation time.  The x-axis shows the value of the stabilization parameter $\theta$. For $\theta = 1$ the quasi-implicit scheme is unconditionally stable. In both cases the stability region for the forward Euler scheme is also shown in blue.}
\end{figure*}

@@

\begin{figure*}
\includegraphics[width=0.9\textwidth]{figures/timestep_convergence_qi.pdf}
\caption{Convergence test with timestep size, quasi-implicit}
\label{fig:timestep_convergence_qi}
\end{figure*}

>>>>

Here we analyze the problem of the sloshing instability.  We demonstrate the nature of the instability, 
as well as clarifying the effect of quasi-implicit stabilization.
The analysis allows us to propose a new time marching scheme for free surface computations with 
good stability properties. It is an explicit method which requires no modification to the system matrix,
which makes it extremely simple to implement.

@@

We begin with the incompressible momentum conservation equations for creeping incompressible flow:
\begin{equation}
\begin{aligned}
\nabla \cdot \mathbf{T} &= \rho \mathbf{g} \\
\nabla \cdot \mathbf{u} &= 0
\end{aligned}
\label{eq:momentum_mass_conservation}
\end{equation}
where $\mathbf{u}$ is the fluid velocity, $\rho$ is the fluid density, and $\mathbf{g}$ is the force due to gravity.
$\mathbf{T}$ is the stress tensor for a Newtonian fluid, defined by
\begin{equation}
\mathbf{T} = 2 \eta \varepsilon(\mathbf{u}) - p \mathbf{I}
\label{eq:stress_tensor}
\end{equation}
where $\eta$ is the viscosity and $\varepsilon(\mathbf{u}) = \frac{1}{2}(\nabla \mathbf{u} + (\nabla \mathbf{u} )^T )$ is the strain-rate tensor.
Substituting the stress tensor into Equation~\eqref{eq:momentum_mass_conservation} gets the familiar form of the Stokes equation:
\begin{equation}
\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p = \rho \mathbf{g}
\label{eq:stokes}
\end{equation}

For the purposes of this analysis it is useful to define a hydrostatic reference state where the 
velocity $\mathbf{u}$ is zero:
\begin{equation}
- \nabla p_0 = \rho_0 \mathbf{g}
\label{eq:hydrostatic_stokes}
\end{equation}
where $p_0$ is the reference hydrostatic pressure and $\rho_0$ is a reference density profile (which may vary with depth).
The total pressure and density can then be decomposed in to variations about their 
reference values: $\rho = \rho_0 + \rho^\prime$, $p = p_0 + p^\prime$.

Furthermore, since we are interested in solving these equations for a domain with a free surface, 
we need an equation to track that surface.
The Lagrangian equation for time evolution of the free surface can be written
\begin{equation}
\frac{d \zeta}{dt} = \mathbf{u \cdot n} \quad \textrm{on  }  \partial \Omega_\textrm{free surface}
\label{eq:surface_evolution}
\end{equation}
where $\zeta$ is the surface coordinate of the free surface, and $\partial \Omega_\textrm{free surface}$ is the 
part of the domain boundary corresponding to the free surface.

@@

In order to better understand the time evolution of Equations~\eqref{eq:momentum_mass_conservation}-\eqref{eq:surface_evolution} 
we will consider the eigenvalues of a simplified homogeneous system. By setting $\rho = \rho_0$, we obtain the system
\begin{equation}
\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p^\prime = 0
\label{eq:homogeneous_stokes}
\end{equation}

\begin{equation}
\nabla \cdot \mathbf{u} = 0
\label{eq:incompressible}
\end{equation}

@@

\begin{equation}
-\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u} ) + \int_{\Omega(t)} p^\prime \nabla \cdot \mathbf{w} 
+ \int_{\partial \Omega_N(t)} \mathbf{w} \cdot \mathbf{T} \cdot \mathbf{n} = 0 
\label{eq:weak_stokes}
\end{equation}

\begin{equation}
\int_{\Omega(t)} q \nabla \cdot \mathbf{u} = 0
\label{eq:weak_incompressible}
\end{equation}
where $\mathbf{w}$ and $q$ are suitably chosen test functions and the integrals over 
$\Omega$ and $\partial \Omega_N$ are over the volume of the domain and the part of the boundary with stress boundary conditions, respectively.
Note that since the free surface can move, the shape of the domain is a function of time.
It is most convenient to integrate over the domain at the current timestep, which corresponds to an explicit scheme.
However, as we will see, the implicit scheme of integrating over the (unknown) domain at the next timestep
is much more stable, at the cost of making the problem nonlinear \citep{furuichi2015implicit}.

The integral over the surface in Equation~\eqref{eq:weak_stokes} accounts for boundary stresses, 
which should be zero when evaluated on a true free surface.
Rather than analyzing finite deformation to the free surface (a nonlinear problem),
we will make the analytically useful approximation of small deformations about the hydrostatic 
reference surface and analyze their stability.
We will therefore  evaluate the integrals in Equation~\eqref{eq:weak_stokes} 
over the hydrostatic reference surface and introduce a temporary auxilliary variable $\zeta$ which 
represents the (small) topography of the free surface relative to that reference surface.
On the reference surface the gravity vector is opposite the direction of the surface normal $\mathbf{g} = -g \mathbf{n}$.
The stress on this reference surface can be approximated by the hydrostatic pressure
\begin{equation}
\mathbf{T} \approx \rho_0 g \zeta \mathbf{I}
\label{eq:hydrostatic}
\end{equation}
Equation~\eqref{eq:weak_stokes} then becomes

\begin{equation}
-\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u} ) + \int_{\Omega(t)} p^\prime \nabla \cdot \mathbf{w} 
+ \int_{\partial \Omega_N(t)} \rho_0 g \zeta  \mathbf{w} \cdot \mathbf{n} = 0 
\end{equation}

We would like to analyze the time evolution of the normal modes of this system: each mode 
is the relaxation of topography with a characteristic relaxation time.  
We denote the normal modes by $\left[ \mathbf{u}_i, p^\prime_i, \zeta_i \right]^T$, with
relaxation times $\tau_i$, where the subscript corresponds to the $i$th normal mode.

@@

\begin{equation}
\frac {d}{d t} \zeta_i(\mathbf{x},t) = \frac{d}{dt} \zeta_i(\mathbf{x})e^{-t/\tau_i} = -\frac{\zeta_i(\mathbf{x},t)}{\tau_i} = \mathbf{u}_i \cdot \mathbf{n}
\end{equation}
This can then be used to eliminate $\zeta$ from the Stokes system:
\begin{equation}
-\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u}_i ) + \int_{\Omega(t)} p^\prime_i \nabla \cdot \mathbf{w} 
= \tau_i \int_{\partial \Omega_N(t)} \rho_0 g (\mathbf{u}_i \cdot \mathbf{n} ) (\mathbf{w} \cdot \mathbf{n})
\label{eq:weak_eigen}
\end{equation}
When these equations are discretized \citep[e.g.][]{kronbichler2012high} we get
\begin{equation}
\begin{bmatrix}
A & B^T \\
B & 0 \\
\end{bmatrix}
\begin{bmatrix}
\mathbf{u}^n_i \\
p^n_i
\end{bmatrix}
=
\tau_i
\begin{bmatrix}
M & 0 \\
0 & 0
\end{bmatrix}
\begin{bmatrix}
\mathbf{u}^n_i \\
p^n_i
\end{bmatrix}
\label{eq:generalized_eigenvalue}
\end{equation}
where $\mathbf{u}^n_i$ and $p^n_i$ are finite-dimensional representations of $\mathbf{u}_i$ and $p^\prime_i$,
and $M$ is the discretization of the bilinear form on the right-hand side of Equation~\eqref{eq:weak_eigen}.

@@

\begin{equation}
\int_{\Omega + \Delta \Omega} \rho  \mathbf{w} \cdot \mathbf{g} \approx
\int_{\Omega} \rho  \mathbf{w} \cdot \mathbf{g} + \theta \Delta t \int_{\partial \Omega} \rho ( \mathbf{w \cdot g}) (\mathbf{u \cdot n} )
\label{eq:kmm}
\end{equation}
The volume integral is the same as that of the unstabilized problem, but now there is a surface integral.
The surface integral has the form of a velocity-dependent surface stress pushing down on the 
domain, and can be thought of as an artificial viscous damping of the surface.
Since it depends on the velocity, it can no longer go on the right-hand side of the 
equation, and must be added as a stabilization term in the system matrix.  
Empirically it has been found to be successful at damping instabilities \citep{kaus2010stabilization, quinquis2011role, duretz2011discretization}.

Indeed, using the formalism in Section~\ref{sec:eigenvalue} we can see the effect of this 
stabilization term. On the reference surface we note that $\mathbf{g} = -g \mathbf{n}$, which
allows us to write the stabilization term as
\begin{equation}
-\theta \Delta t \int_{\partial \Omega} \rho g ( \mathbf{w \cdot n}) (\mathbf{u \cdot n} )
\end{equation}
The integral here is precisely the same as that on the right hand side of Equation~\eqref{eq:weak_eigen}, which was discretized as the matrix $M$.

@@

\subsection{Stability of the scheme}
The parameter $\tau^*$ sets how quickly $\mathbf{u}$ decays in Equation~\eqref{eq:velocity_decay}, and a good 
choice is crucial for accuracy and stability. A shorter decay time corresponds to more stabilization,
but if it becomes too short, the surface velocity will become overdamped. In general, we will want 
to choose $\tau^*$ so that it is as close as possible to the relaxation time of the least stable mode, or $\tau_{\mathrm{min}}$.

@@

The stability of this scheme is determined by the choice of $\tau^*$.  
It has a region of unconditional stability, where
\begin{equation}
\tau^* \le 2 \tau_i
\end{equation}
Otherwise, the scheme is conditionally stable, with the timestep limited by 
\begin{equation}
\Delta t \le -\tau^* \log \left(1 - 2 \frac{\tau_i}{\tau^*} \right)
\end{equation}
The stability region is plotted in Figure \ref{fig:stability_region}.

\begin{figure*}
\includegraphics[width=0.9\textwidth]{figures/stability_region.pdf}
\label{fig:stability_region}
\caption{Left: stability region for the nonstandard finite difference scheme (green). On the x-axis is the value of the stabilization time, in units of the minimum relaxation time $\tau_{\mathrm{min}}$.  On the y-axis is the value of the timestep, also in units of the minimum relaxation time. For $\tau^*\le 2 \tau_{\mathrm{min}}$ the nonstandard finite difference scheme is unconditionally stable.  Right: stability region for the quasi-implicit scheme (green).  Again, the y axis is in units of the minumum relaxation time.  The x-axis shows the value of the stabilization parameter $\theta$. For $\theta = 1$ the quasi-implicit scheme is unconditionally stable. In both cases the stability region for the forward Euler scheme is also shown in blue.}
\end{figure*}

@@
